\section{Research Motivation}

This research is primarily motivated by the need to enable autonomous agricultural robots to perform complex manipulation tasks, specifically robotic pruning. Unlike simple navigation, pruning requires a high-fidelity localisation and mapping (SLAM) framework that can provide the millimetre-level precision necessary for a robotic arm to interact with thin, overlapping vine structures. Developing a system that can maintain this accuracy in the presence of dynamic foliage and repetitive, unstructured row patterns is a critical bottleneck in automating viticulture and horticulture.

The demand for such technology is underscored by the economic landscape of New Zealand's primary industries. As of 2026, horticulture export revenue is forecast to reach a record \$9.2 billion, with wine production constituting \$2.3 billion \cite{industriesSituationOutlookPrimary2025}. However, the sector faces chronic labour shortages and rising operational costs, with manual pruning alone accounting for around 23\% of annual vineyard expenses \cite{2025MarlboroughVineyard}. The 2025 Vineyard Monitoring Report published by the Ministry for Primary Industries indicates that during the 2024/2025 season, expenditure on manual pruning increased, while spending on machine pruning declined by 21\%. This shift is attributed to grower concerns regarding vine damage caused by current mechanical stripping machines \cite{2025MarlboroughVineyard}. These concerns highlight a critical limitation of current commercial machine-based pruning systems: insufficient precision to meet vineyard managers' operational requirements. Overcoming these precision challenges would enable robotic labour to provide a scalable response to ongoing workforce constraints, thereby supporting the long-term sustainability of New Zealand’s high-value horticultural crops.

Beyond the immediate application to pruning, this work seeks to advance the research objectives of the UCVision Research Lab. By improving SLAM robustness in vegetated environments, this thesis contributes to the lab's broader mission of "seeing through occlusion" and tracking plant growth across multiple stages. The frameworks developed for semi-structured orchard conditions extend directly to other challenging environments such as forestry or search-and-rescue, providing a foundation for autonomous systems in visually complex terrain.




