\section{Research Context}

This thesis is situated within the broader research initiatives of the University of Canterbury Computer Vision Research Group (UCVision), specifically contributing to 'The Occlusion Project'. This project utilises autonomous field robots to capture multi-view imagery of vineyards and orchards, which are subsequently processed \textit{offline} to generate high-fidelity 3D reconstructions using 3D Gaussian Splatting (3DGS). These models serve as the "digital twin" required for complex agricultural tasks such as precision yield estimation and automated labour.

\vspace{\medskipamount}

A primary objective of this project is the automation of vineyard pruning. In this workflow, pruning decisions and robotic arm path planning are computed within the high-fidelity 3DGS environment. Unlike conventional pruning systems that rely on lower-resolution, real-time sensor data for immediate modelling and execution, the UCVision workflow decouples data collection from intervention. While this allows for superior structural analysis of the vine, it introduces a critical spatial synchronisation problem: the realignment of global and local coordinate frames.

\vspace{\medskipamount}

For the robot to execute a pre-computed pruning plan, it must bridge the gap between its real-time physical state and the coordinate system of the offline model. This necessitates a robust SLAM framework capable of solving three distinct sub-problems:
\begin{enumerate}
    \item \textbf{Global Relocalization:} Navigating the platform back to the specific plant or row identified in the modeling phase.
    \item \textbf{Strategic Positioning:} Aligning the mobile base within the physical workspace constraints of the robotic arm to ensure all required cutpoints are reachable.
    \item \textbf{Sim-to-Real Registration:} Achieving millimeter-level alignment between the robot’s onboard perception and the 3D model. This ensures that the simulated path-planner’s coordinates map precisely to the physical vine structures in real life.
\end{enumerate}

\vspace{\medskipamount}

This thesis addresses this capability gap by developing a SLAM and relocalisation framework that facilitates this alignment. By synchronizing the robot’s perceived position with both the physical environment and the pre-existing 3D model, this research enables the transition from static digital twins to active, autonomous robotic intervention in the field.